VOI-aware Monte Carlo Sampling in Trees

### 2011-12-04 paper/vago/paper.tex
paper in progress

\begin{abstract}
Upper bounds for the VOI are provided for pure exploration in the
Multi-armed Bandit Problem. Sampling policies based on the upper
bounds are suggested. Empirical evaluation of the policies is provided
on random problems as well on the Go game.
\end{abstract}

@@

\begin{thm} The blinkered estimate $\Lambda_i^b$ of intrinsic value of information of sampling
  the $i$th arm for the remaining budget of $N$ samples is bounded from above as
\begin{equation}
  \Lambda_i^b \le \left\{
  \begin{array}{l l}
    \Pr(\overline X_i'\le\overline X_\beta)\overline X_\beta\frac N
       {N+n_i} < \Pr(\overline X_i'\le\overline X_\beta)\overline X_\beta\frac N
       {n_i} & \mbox{if $i=\alpha$} \\
    \Pr(\overline X_i'\ge\overline X_\alpha)(1-\overline X_\alpha)\frac N {N+n_i}&  \mbox{otherwise}
  \end{array} \right.
\label{eqn:thm-be}
\end{equation}
where $\overline X_i'$ is the sample mean of the $i$th arm after $N$ more
samples.
\end{thm}

 The probabilities in equations (\ref{eqn:thm-vopi}, \ref{eqn:thm-be}) can be bounded from above using concentration
inequalities. In particular, Lemma~\ref{lemma:hoeffding-prob-bounds} is
based on the Hoeffding inequality (\ref{eqn:conc-hoeffding}):
\begin{lmm} The probabilities in equations (\ref{eqn:thm-vopi}, \ref{eqn:thm-be}) are bounded from above as
\begin{eqnarray}
\Pr(\IE[X_i] \le \overline X_\beta|i=\alpha)& \le & \exp(-2 (\overline X_i - \overline X_\beta)^2 n_i)\nonumber\\
\Pr(\IE[X_i] \ge \overline X_\alpha|i\ne\alpha)& \le & \exp(-2 (\overline X_\alpha - \overline X_i)^2 n_i)\nonumber\\
\Pr(X_i' \le \overline X_\beta|i=\alpha)& \le & 2\exp(-0.5(\overline X_i - \overline X_\beta)^2 n_i)\nonumber\\
\Pr(X_i' \ge \overline X_\alpha|i\ne\alpha)& \le & 2\exp(-0.5(\overline X_\alpha - \overline X_i)^2 n_i)
\label{eqn:probound-perf-hoeffding}
\end{eqnarray}
\label{lemma:hoeffding-prob-bounds}
\end{lmm}

Better bounds can be obtained through tighter estimates on
the probabilities, for example, based on the empirical Bernstein
inequality (\ref{eqn:conc-empbernstein}).

@@

@@

>>>>

\begin{abstract}
Upper bounds for the VOI are provided for pure exploration in the
Multi-armed Bandit Problem. Sampling policies based on the upper
bounds are suggested. Empirical evaluation of the policies is provided
on random problems as well as on the Go game.
\end{abstract}

@@

\begin{thm} The blinkered estimate $\Lambda_i^b$ of intrinsic value of information of sampling
  the $i$th arm for the remaining budget of $N$ samples is bounded from above as
\begin{equation}
  \Lambda_i^b \le \left\{
  \begin{array}{l l}
    \Pr(\overline X_i'\le\overline X_\beta)\overline X_\beta\frac N {N+n_i} < \Pr(\overline X_i'\le\overline X_\beta)\overline X_\beta\frac N   {n_i} & \mbox{if $i=\alpha$} \\
    \Pr(\overline X_i'\ge\overline X_\alpha)(1-\overline X_\alpha)\frac N {N+n_i} < \Pr(\overline X_i'\ge\overline X_\alpha)(1-\overline X_\alpha)\frac N {n_i}&  \mbox{otherwise}
  \end{array} \right.
\label{eqn:thm-be}
\end{equation}
where $\overline X_i'$ is the sample mean of the $i$th arm after $N$ more
samples.
\end{thm}

 The probabilities in equations (\ref{eqn:thm-vopi}, \ref{eqn:thm-be}) can be bounded from above using concentration
inequalities. In particular, Lemma~\ref{lemma:hoeffding-prob-bounds} is
based on the Hoeffding inequality (\ref{eqn:conc-hoeffding}):
\begin{lmm} The probabilities in equations (\ref{eqn:thm-vopi}, \ref{eqn:thm-be}) are bounded from above as
\begin{eqnarray}
\Pr(\IE[X_i] \le \overline X_\beta|i=\alpha)& \le & \exp(-2 (\overline X_i - \overline X_\beta)^2 n_i)\nonumber\\
\Pr(\IE[X_i] \ge \overline X_\alpha|i\ne\alpha)& \le & \exp(-2 (\overline X_\alpha - \overline X_i)^2 n_i)\nonumber\\
\Pr(X_i' \le \overline X_\beta|i=\alpha)& \le & 2\exp\left(-\frac {(\overline X_i - \overline X_\beta)^2 n_i} 2 \right)\nonumber\\
\Pr(X_i' \ge \overline X_\alpha|i\ne\alpha)& \le & 2\exp\left(-\frac {(\overline X_\alpha - \overline X_i)^2 n_i} 2 \right)
\label{eqn:probound-perf-hoeffding}
\end{eqnarray}
\label{lemma:hoeffding-prob-bounds}
\end{lmm}

Better bounds can be obtained through tighter estimates on
the probabilities, for example, based on the empirical Bernstein
inequality (\ref{eqn:conc-empbernstein}) or through a more careful
application of the Hoeffding inequality (Appendix~\ref{app:better-hoeffding-bound}).

@@

\section{VOI-based Sampling Control}

\subsection{Selection Criterion}

Following the principles of rational metareasoning, an arm with the highest upper bound $\hat
\Lambda$ on the perfect value of information should be pulled at each
step. This way, arms known to have a low VOI  would be pulled less frequently.

\subsection{Termination Condition}

The upper bounds~(\ref{eq:lambda-hoeffding-bounds}, \ref{eq:lambda-bernstein-bounds}) decrease exponentially with the
number of pulls $n$. When the upper bound of the VOI for all arms
becomes lower than a threshold $\lambda$, which can be chosen based on
resource constraints, the sampling should be stopped, and an arm should
be chosen.

\subsection{Sample Redistribution in Trees}

UCT forwards samples to the next search stage when the winner at the
current state is known with high confidence. VOI-based termination
condition can be used to stop the sampling in the current state early
and save the remaining samples for re-use in a later search state.

@@

\section{Better Hoeffding-Based Bound on Value of Perfect Information}
\label{app:better-hoeffding-bound}

The bound can be supposedly be improved by selecting a midpoint
$0 < \gamma < \overline X_\beta$ and computing the bound as the sum of two parts:
\begin{itemize}
\item $\overline X_\beta-\gamma$ multiplied by the probability that
  $\mu_\alpha \le \overline X_\beta$;
\item $\overline X_\beta$ multiplied by the probability that $\mu_\alpha\le
  \gamma$.
\end{itemize}.
\[ V_{\overline X=\overline X_\alpha} \le (\overline X_\beta-\gamma)\exp\left(-2n(\overline X_\alpha-\overline X_\beta)^2\right)+\overline X_\beta\exp\left(-2n(\overline X_\alpha-\gamma)^2\right) \]

\vspace{\baselineskip}

The minimum of $V^*$ is achieved when $\frac {dV^*} {d\gamma}=0$, that
is, when $\gamma$ is the root of the following equation:
\[ 4\overline X_\beta n(\overline X_\alpha-\gamma)=\exp\left(-2n\left(\frac {\overline X_\alpha-\overline X_\beta}
    {\overline X_\alpha-\gamma}\right)^2\right) \]
If a root in the interval  $0\le\gamma\le\beta$ exists, then the
number of samples is bounded as
\[n\le\frac 1 {4\overline X_\beta(\overline X_\alpha-\overline X_\beta)}\]
by observing that the right-hand side 
is at most $1$ (a negative power), and the left-hand side is at least
$4\overline X_\beta n(\overline X_\alpha-\overline X_\beta)$.
So, the bound can supposedly be improved for smaller values of
$n$. The improvement is more significant when the current best and
second-best sample means are close.

The derivation for the other case
(sampling an item that can be better than the current best) is
obtained by substitution $1-\overline X, 1-\overline X_\alpha, 1-\gamma$ instead of
$\overline X_\alpha, \overline X_\beta, \gamma$:

\[ V_{\overline X\ne\overline X_\alpha} \le (\gamma-\overline X_\alpha)\exp\left(-2n(\overline X_\alpha-\overline
  X)^2\right)+(1-\overline X_\alpha)\exp\left(-2n(\gamma-\overline X)^2\right) \]

The anticipated influence of the improved
VOI estimate would be that the selected item will be a less discovered
one and further from the current best or second-best.

A closed-form solution for $\gamma$ cannot be obtained, but given
$\overline X_\alpha, \overline X_\beta, n$, the value of $\gamma$ can be efficiently
computed. It should be determined empirically whether the improved
estimate has justified influence on the performance of the algorithm.
